taxize: taxonomic search and retrieval in R

### 2013-09-30 f1000_rev/taxize_ms.tex
added nfew version of ms to f1000_rev folder, folded in changes from reviewers, issue #178

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Welcome to writeLaTeX --- just edit your LaTeX on the left,
% and we'll compile it for you on the right. If you give 
% someone the link to this page, they can edit at the same
% time. See the help menu above for more info. Enjoy!
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% For more detailed article preparation guidelines, please see:
% http://f1000research.com/author-guidelines

\documentclass[10pt,a4paper,twocolumn]{article}
\usepackage{f1000_styles}
\usepackage{hyperref}
\usepackage{listings}

@@

\usepackage{listings}

@@

\begin{abstract}

All species are hierarchically related to one another, and we use taxonomic names to label the nodes in this hierarchy. Taxonomic data is becoming increasingly available on the web, but scientists need a way to access it in a programmatic fashion that's easy and reproducible. We have developed taxize, an open-source software package (freely available from \url{http://cran.r-project.org/web/packages/taxize/index.html}) for the R language. taxize provides simple, programmatic access to taxonomic data for 13 data sources around the web. We discuss the need for a taxonomic toolbelt in R, and outline a suite of use cases for which taxize is ideally suited (including a full workflow as an appendix). The taxize package facilitates open and reproducible science by allowing taxonomic data collection to be done in the open-source R platform.

\end{abstract}
\clearpage

\section*{Introduction}

Evolution by natural selection has led to a hierarchical relationship among all living organisms.  Thus, species are categorized using a taxonomic hierarchy, starting with the binomial species name (e.g, \emph{Homo sapiens}), moving up to genus (\emph{Homo}), then family (\emph{Hominidae}), and on up to Domain (\emph{Eukarya}). Biologists, whether studying organisms at the cell, organismal, or community level, can put their study objects into taxonomic context, allowing them to know close and distant relatives, find relevant literature, and more.

@@

\section*{Data sources and package details}

taxize uses many data sources (Table~\ref{tab:a}), and more can easily be added. There are two common tasks provided by the data sources: name search and name resolution. Other functionality in taxize includes retrieving a classification tree for a species, or retrieving child taxa of a focal taxon. One of the data sources (Phylomatic) returns phylogenies, while another (NCBI) returns genetic sequence data. However, there are other R packages that are focused solely on sequence data, such as rsnps \cite{chamberlain2013}, rentrez \cite{winter2013}, BoSSA \cite{lefeuvre2010}, and ape \cite{paradis2004}, so taxize does not venture deeply into these other domains.

@@

One available data source in taxize is The Plant List \url{http://www.theplantlist.org}. The connection in taxize is done via the \emph{taxonstand} package \cite{Cayuela_2012} that solely interacts with the Plant List. We provide a few convenience functions that wrap \emph{taxonstand} into taxize.

@@

\begin{knitrout}
\begin{scriptsize}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{install.packages}(\hlstring{"taxize"})
\hlfunctioncall{library}(taxize)
\end{alltt}
\end{kframe}
\end{scriptsize}
\end{knitrout}

Advanced users can also download and install the latest development copy from GitHub \url{https://github.com/ropensci/taxize_}.

@@

\begin{knitrout}
\begin{scriptsize}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
temp <- \hlfunctioncall{gnr_resolve}(names = \hlfunctioncall{c}(\hlstring{"Helianthos annus"}, \newline \hlstring{"Homo saapiens"}))
temp[, -\hlfunctioncall{c}(1, 4)]
\end{alltt}
\begin{verbatim}
                 matched_name       data_source_title
1        Helianthus annuus L.       Catalogue of Life
2            Helianthus annus GBIF Taxonomic Backbone
3            Helianthus annus                     EOL
4         Helianthus annus L.                     EOL
5            Helianthus annus           uBio NameBank
6 Homo sapiens Linnaeus, 1758       Catalogue of Life
\end{verbatim}
\end{kframe}
\end{scriptsize}
\end{knitrout}

@@

\begin{knitrout}
\begin{scriptsize}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
mynames <- \hlfunctioncall{c}(\hlstring{"Helianthus annuus"}, \hlstring{"Pinus contort"}, \hlstring{"Poa anua"}, \newline \hlstring{"Abis magnifica"}, \hlstring{"Rosa california"}, \hlstring{"Festuca arundinace"}, \newline \hlstring{"Sorbus occidentalos"}, \hlstring{"Madia sateva"})
\hlfunctioncall{tnrs}(query = mynames)[, -\hlfunctioncall{c}(5:7)]
\end{alltt}
\begin{verbatim}
         submittedName        acceptedName    sourceId score
9    Helianthus annuus   Helianthus annuus iPlant_TNRS     1
10   Helianthus annuus   Helianthus annuus        NCBI     1
4        Pinus contort      Pinus contorta iPlant_TNRS  0.98
5             Poa anua           Poa annua iPlant_TNRS  0.96
3       Abis magnifica     Abies magnifica iPlant_TNRS  0.96
7      Rosa california    Rosa californica iPlant_TNRS  0.99
8      Rosa california          California        NCBI     1
2   Festuca arundinace Festuca arundinacea iPlant_TNRS  0.99
1  Sorbus occidentalos Sorbus occidentalis iPlant_TNRS  0.99
6         Madia sateva        Madia sativa iPlant_TNRS  0.97
\end{verbatim}
\end{kframe}
\end{scriptsize}
\end{knitrout}

@@

Another common use case is when there are many synonyms for a species. In this example, we have three synonyms of the currently accepted name for a species.

\begin{knitrout}
\begin{scriptsize}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
mynames <- \hlfunctioncall{c}(\hlstring{"Helianthus annuus ssp. jaegeri"}, \newline \hlstring{"Helianthus annuus ssp. lenticularis"}, \newline \hlstring{"Helianthus annuus ssp. texanus"}, \newline \hlstring{"Helianthus annuus var. lenticularis"}, \newline \hlstring{"Helianthus annuus var. macrocarpus"}, \newline \hlstring{"Helianthus annuus var. texanus"})
tsn <- \hlfunctioncall{get_tsn}(mynames)
\hlfunctioncall{ldply}(tsn, itis_acceptname)
\end{alltt}
\begin{verbatim}
  submittedTsn      acceptedName acceptedTsn
1       525928 Helianthus annuus       36616
2       525929 Helianthus annuus       36616
3       525930 Helianthus annuus       36616
4       536095 Helianthus annuus       36616
5       536096 Helianthus annuus       36616
6       536097 Helianthus annuus       36616
\end{verbatim}
\end{kframe}
\end{scriptsize}
\end{knitrout}

@@

\begin{knitrout}
\begin{scriptsize}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
specieslist <- \hlfunctioncall{c}(\hlstring{"Abies procera"}, \hlstring{"Pinus contorta"})
\hlfunctioncall{classification}(specieslist, db = \hlstring{"itis"})
\end{alltt}
\begin{verbatim}
$`Abies procera`
        rankName       taxonName    tsn
1        Kingdom         Plantae 202422
2     Subkingdom  Viridaeplantae 846492
3   Infrakingdom    Streptophyta 846494
4       Division    Tracheophyta 846496
5    Subdivision Spermatophytina 846504
6  Infradivision    Gymnospermae 846506
7          Class       Pinopsida 500009
8          Order         Pinales 500028
9         Family        Pinaceae  18030
10         Genus           Abies  18031
11       Species   Abies procera 181835

$`Pinus contorta`
        rankName       taxonName    tsn
1        Kingdom         Plantae 202422
2     Subkingdom  Viridaeplantae 846492
3   Infrakingdom    Streptophyta 846494
4       Division    Tracheophyta 846496
5    Subdivision Spermatophytina 846504
6  Infradivision    Gymnospermae 846506
7          Class       Pinopsida 500009
8          Order         Pinales 500028
9         Family        Pinaceae  18030
10         Genus           Pinus  18035
11       Species  Pinus contorta 183327
\end{verbatim}
\end{kframe}
\end{scriptsize}
\end{knitrout}

@@

\begin{knitrout}
\begin{scriptsize}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{tax_name}(query = \hlstring{"Helianthus annuus"}, get = \hlstring{"family"}, \newline db = \hlstring{"itis"})
\end{alltt}
\begin{verbatim}
      family
1 Asteraceae
\end{verbatim}
\begin{alltt}
\hlfunctioncall{tax_name}(query = \hlstring{"Helianthus annuus"}, get = \hlstring{"family"}, \newline db = \hlstring{"ncbi"})
\end{alltt}
\begin{verbatim}
      family
1 Asteraceae
\end{verbatim}
\end{kframe}
\end{scriptsize}
\end{knitrout}

@@

\begin{knitrout}
\begin{scriptsize}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{get_tsn}(searchterm = \hlstring{"Heliastes"}, searchtype = \hlstring{"sciname"})
\end{alltt}
\begin{verbatim}
           combinedname    tsn
1     Heliastes bicolor 615238
2   Heliastes chrysurus 615250
3     Heliastes cinctus 615573
4  Heliastes dimidiatus 615257
5  Heliastes hypsilepis 615273
6 Heliastes immaculatus 615639
7 Heliastes opercularis 615300
8      Heliastes ovalis 615301
 1 
NA 
attr(,"class")
[1] "tsn"
\end{verbatim}
\end{kframe}
\end{scriptsize}
\end{knitrout}

@@

\begin{knitrout}
\begin{scriptsize}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
splist <- \hlfunctioncall{c}(\hlstring{"annona cherimola"}, \hlstring{"annona muricata"}, \newline \hlstring{"quercus robur"}, \hlstring{"shorea robusta"}, \hlstring{"pandanus patina"}, \newline \hlstring{"oryza sativa"}, \hlstring{"durio zibethinus"})
\hlfunctioncall{get_tsn}(searchterm = splist, searchtype = \hlstring{"sciname"})
\end{alltt}
\begin{verbatim}
[1] "506198" "18098"  "19405"  "506787" "507376" "41976"  "506099"
attr(,"class")
[1] "tsn"
\end{verbatim}
\end{kframe}
\end{scriptsize}
\end{knitrout}

@@

\begin{knitrout}
\begin{scriptsize}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{get_uid}(sciname = \hlfunctioncall{c}(\hlstring{"Chironomus riparius"}, \hlstring{"aaa vva"}))
\end{alltt}
\begin{verbatim}
[1] "315576" NA      
attr(,"class")
[1] "uid"
\end{verbatim}
\end{kframe}
\end{scriptsize}
\end{knitrout}

\subsection*{Retrieve a phylogeny}
Ecologists are increasingly taking a phylogenetic approach to ecology, applying phylogenies to topics such as the study of community structure \cite{webb2002phylogenies}, ecological networks \cite{rafferty2013phylogenetic}, and functional trait ecology \cite{poff2006functional}. Yet, many biologists are not adequately trained in reconstructing phylogenies. Fortunately, there are some sources for getting a phylogeny without having to know how to build one; one of these is for angiosperms, called Phylomatic \cite{webb2005}. We have created a workflow in taxize that accepts a species list, and taxize works behind the scenes to get higher taxonomic names, which are required by Phylomatic to get a phylogeny. Here is a short example, producing the tree in Fig. \ref{fig:phylomatic}.

\begin{knitrout}
\begin{scriptsize}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
taxa <- \hlfunctioncall{c}(\hlstring{"Poa annua"}, \hlstring{"Abies procera"}, \hlstring{"Helianthus annuus"})
tree <- \hlfunctioncall{phylomatic_tree}(taxa = taxa)
tree$tip.label <- \hlfunctioncall{capwords}(tree$tip.label)
\hlfunctioncall{plot}(tree, cex = 1)
\end{alltt}
\end{kframe}
\end{scriptsize}
\end{knitrout}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=3in]{phylomaticphylogeny.pdf}
\end{center}
\caption{
{A phylogeny for three species. This phylogeny was produced using the \emph{phylomatic\_tree} function, which queries the Phylomatic database, and prunes a previously created phylogeny of plants.}
}
\label{fig:phylomatic}
\end{figure}

@@

\subsection*{What taxa are the children of my taxon of interest?}

If someone is not a taxonomic specialist on a particular taxon they probably do not know what children taxa are within a family, or within a genus. This task becomes especially unwieldy when there are a large number of taxa downstream. You can of course go to a website like Wikispecies \url{http://species.wikimedia.org/wiki/Main_Page} or Encyclopedia of Life \url{http://eol.org/} to get downstream names. However, taxize provides an easy way to programatically search for downstream taxa, both for the Catalogue of Life (CoL) \url{http://www.catalogueoflife.org/} and the Integrated Taxonomic Information System \url{http://www.itis.gov/}. Here is a short example using the CoL in which we want to find all the species within the genus \emph{Apis} (honey bees).

\begin{knitrout}
\begin{scriptsize}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{col_downstream}(name = \hlstring{"Apis"}, downto = \hlstring{"Species"})[[1]]
\end{alltt}
\begin{verbatim}
  childtaxa_id     childtaxa_name childtaxa_rank
1      6971712 Apis andreniformis        Species
2      6971713        Apis cerana        Species
3      6971714       Apis dorsata        Species
4      6971715        Apis florea        Species
5      6971716 Apis koschevnikovi        Species
6      6845885     Apis mellifera        Species
7      6971717   Apis nigrocincta        Species
\end{verbatim}
\end{kframe}
\end{scriptsize}
\end{knitrout}

@@

\subsection*{Matching species tables with different taxonomic resolution}

Biologists often need to match different sets of data tied to species. For example, trait-based approaches are a promising tool in ecology \cite{statzner_can_2010}. One problem is that abundance data must be matched with trait databases such as the NCBI Taxonomy database \cite{usseglio-polatera_biological_2000}. These two data tables may contain species information on different taxonomic levels and data might have to be aggregated to a joint taxonomic level, so that the data can be merged. taxize can help in this data-cleaning step, providing a reproducible workflow.

>>>>

\documentclass[10pt,a4paper,twocolumn]{article}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\usepackage{f1000_styles}
\usepackage{hyperref}
\usepackage{listings}

@@

\usepackage{listings}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}

@@

\begin{abstract}
All species are hierarchically related to one another, and we use taxonomic names to label the nodes in this hierarchy. Taxonomic data is becoming increasingly available on the web, but scientists need a way to access it in a programmatic fashion that's easy and reproducible. We have developed taxize, an open-source software package (freely available from \url{http://cran.r-project.org/web/packages/taxize/index.html}) for the R language. taxize provides simple, programmatic access to taxonomic data for 13 data sources around the web. We discuss the need for a taxonomic toolbelt in R, and outline a suite of use cases for which taxize is ideally suited (including a full workflow as an appendix). The taxize package facilitates open and reproducible science by allowing taxonomic data collection to be done in the open-source R platform.
\end{abstract}
\clearpage

\section*{Introduction}
Evolution by natural selection has led to a hierarchical relationship among all living organisms.  Thus, species are categorized using a taxonomic hierarchy, starting with the binomial species name (e.g, \emph{Homo sapiens}), moving up to genus (\emph{Homo}), then family (\emph{Hominidae}), and on up to Domain (\emph{Eukarya}). Biologists, whether studying organisms at the cell, organismal, or community level, can put their study objects into taxonomic context, allowing them to know close and distant relatives, find relevant literature, and more.

@@

\section*{Data sources and package details}
taxize uses many data sources (Table~\ref{tab:a}), and more can easily be added. There are two common tasks provided by the data sources: name search and name resolution. Other functionality in taxize includes retrieving a classification tree for a species, or retrieving child taxa of a focal taxon. One of the data sources (Phylomatic) returns phylogenies, while another (NCBI) returns genetic sequence data. However, there are other R packages that are focused solely on sequence data, such as rsnps \cite{chamberlain2013}, rentrez \cite{winter2013}, BoSSA \cite{lefeuvre2010}, and ape \cite{paradis2004}, so taxize does not venture deeply into these other domains.

@@

One available data source in taxize is The Plant List \url{http://www.theplantlist.org}. The connection in taxize is done via the \emph{taxonstand} package \cite{Cayuela_2012,taxonstand} that solely interacts with the Plant List. We provide a few convenience functions that wrap \emph{taxonstand} into taxize.

@@

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{install.packages}\hlstd{(}\hlstr{"taxize"}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{library}\hlstd{(taxize)}
\end{alltt}
\end{kframe}
\end{knitrout}

Advanced users can also download and install the latest development copy from GitHub \url{https://github.com/ropensci/taxize_}, also permanently available at \url{http://dx.doi.org/10.5281/zenodo.7097}.

@@

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{temp} \hlkwb{<-} \hlkwd{gnr_resolve}\hlstd{(}\hlkwc{names} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"Helianthos annus"}\hlstd{,} \hlstr{"Homo saapiens"}\hlstd{))}
\hlstd{temp[,} \hlopt{-}\hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{,} \hlnum{4}\hlstd{)]}
\end{alltt}
\end{kframe}
\end{knitrout}

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
                 matched_name       data_source_title
1        Helianthus annuus L.       Catalogue of Life
2            Helianthus annus GBIF Taxonomic Backbone
3            Helianthus annus                     EOL
4         Helianthus annus L.                     EOL
5            Helianthus annus           uBio NameBank
6 Homo sapiens Linnaeus, 1758       Catalogue of Life
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{mynames} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"Helianthus annuus"}\hlstd{,} \hlstr{"Pinus contort"}\hlstd{,} \hlstr{"Poa anua"}\hlstd{,} \hlstr{"Abis magnifica"}\hlstd{,}
    \hlstr{"Rosa california"}\hlstd{,} \hlstr{"Festuca arundinace"}\hlstd{,} \hlstr{"Sorbus occidentalos"}\hlstd{,} \hlstr{"Madia sateva"}\hlstd{)}
\hlkwd{tnrs}\hlstd{(}\hlkwc{query} \hlstd{= mynames)[,} \hlopt{-}\hlkwd{c}\hlstd{(}\hlnum{5}\hlopt{:}\hlnum{7}\hlstd{)]}
\end{alltt}
\end{kframe}
\end{knitrout}

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
         submittedName        acceptedName    sourceId score
9    Helianthus annuus   Helianthus annuus iPlant_TNRS     1
10   Helianthus annuus   Helianthus annuus        NCBI     1
4        Pinus contort      Pinus contorta iPlant_TNRS  0.98
5             Poa anua           Poa annua iPlant_TNRS  0.96
3       Abis magnifica     Abies magnifica iPlant_TNRS  0.96
7      Rosa california    Rosa californica iPlant_TNRS  0.99
8      Rosa california          California        NCBI     1
2   Festuca arundinace Festuca arundinacea iPlant_TNRS  0.99
1  Sorbus occidentalos Sorbus occidentalis iPlant_TNRS  0.99
6         Madia sateva        Madia sativa iPlant_TNRS  0.97
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

Another common use case is when there are many synonyms for a species. In this example, we have six synonyms of the currently accepted name for a species.

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{library}\hlstd{(plyr)}
\hlstd{mynames} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"Helianthus annuus ssp. jaegeri"}\hlstd{,} \hlstr{"Helianthus annuus ssp. lenticularis"}\hlstd{,}
    \hlstr{"Helianthus annuus ssp. texanus"}\hlstd{,} \hlstr{"Helianthus annuus var. lenticularis"}\hlstd{,}
    \hlstr{"Helianthus annuus var. macrocarpus"}\hlstd{,} \hlstr{"Helianthus annuus var. texanus"}\hlstd{)}
\hlstd{tsn} \hlkwb{<-} \hlkwd{get_tsn}\hlstd{(mynames)}
\hlkwd{ldply}\hlstd{(tsn, itis_acceptname)}
\end{alltt}
\end{kframe}
\end{knitrout}

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
  submittedTsn      acceptedName acceptedTsn
1       525928 Helianthus annuus       36616
2       525929 Helianthus annuus       36616
3       525930 Helianthus annuus       36616
4       536095 Helianthus annuus       36616
5       536096 Helianthus annuus       36616
6       536097 Helianthus annuus       36616
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{specieslist} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"Abies procera"}\hlstd{,} \hlstr{"Pinus contorta"}\hlstd{)}
\hlkwd{classification}\hlstd{(specieslist,} \hlkwc{db} \hlstd{=} \hlstr{"itis"}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
$`Abies procera`
        rankName       taxonName    tsn
1        Kingdom         Plantae 202422
2     Subkingdom  Viridaeplantae 846492
3   Infrakingdom    Streptophyta 846494
4       Division    Tracheophyta 846496
5    Subdivision Spermatophytina 846504
6  Infradivision    Gymnospermae 846506
7          Class       Pinopsida 500009
8          Order         Pinales 500028
9         Family        Pinaceae  18030
10         Genus           Abies  18031
11       Species   Abies procera 181835

$`Pinus contorta`
        rankName       taxonName    tsn
1        Kingdom         Plantae 202422
2     Subkingdom  Viridaeplantae 846492
3   Infrakingdom    Streptophyta 846494
4       Division    Tracheophyta 846496
5    Subdivision Spermatophytina 846504
6  Infradivision    Gymnospermae 846506
7          Class       Pinopsida 500009
8          Order         Pinales 500028
9         Family        Pinaceae  18030
10         Genus           Pinus  18035
11       Species  Pinus contorta 183327
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{tax_name}\hlstd{(}\hlkwc{query} \hlstd{=} \hlstr{"Helianthus annuus"}\hlstd{,} \hlkwc{get} \hlstd{=} \hlstr{"family"}\hlstd{,} \hlkwc{db} \hlstd{=} \hlstr{"itis"}\hlstd{)}
\hlkwd{tax_name}\hlstd{(}\hlkwc{query} \hlstd{=} \hlstr{"Helianthus annuus"}\hlstd{,} \hlkwc{get} \hlstd{=} \hlstr{"family"}\hlstd{,} \hlkwc{db} \hlstd{=} \hlstr{"ncbi"}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
      family
1 Asteraceae
      family
1 Asteraceae
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{get_tsn}\hlstd{(}\hlkwc{searchterm} \hlstd{=} \hlstr{"Heliastes"}\hlstd{,} \hlkwc{searchtype} \hlstd{=} \hlstr{"sciname"}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
           combinedname    tsn
1     Heliastes bicolor 615238
2   Heliastes chrysurus 615250
3     Heliastes cinctus 615573
4  Heliastes dimidiatus 615257
5  Heliastes hypsilepis 615273
6 Heliastes immaculatus 615639
7 Heliastes opercularis 615300
8      Heliastes ovalis 615301
       1 
"615238" 
attr(,"class")
[1] "tsn"
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{splist} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"annona cherimola"}\hlstd{,} \hlstr{"annona muricata"}\hlstd{,} \hlstr{"quercus robur"}\hlstd{,} \hlstr{"shorea robusta"}\hlstd{,}
    \hlstr{"pandanus patina"}\hlstd{,} \hlstr{"oryza sativa"}\hlstd{,} \hlstr{"durio zibethinus"}\hlstd{)}
\hlkwd{get_tsn}\hlstd{(}\hlkwc{searchterm} \hlstd{= splist,} \hlkwc{searchtype} \hlstd{=} \hlstr{"sciname"}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
[1] "506198" "18098"  "19405"  "506787" "507376" "41976"  "506099"
attr(,"class")
[1] "tsn"
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{get_uid}\hlstd{(}\hlkwc{sciname} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"Chironomus riparius"}\hlstd{,} \hlstr{"aaa vva"}\hlstd{))}
\end{alltt}
\end{kframe}
\end{knitrout}

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
[1] "315576" NA      
attr(,"class")
[1] "uid"
\end{verbatim}
\end{kframe}
\end{knitrout}

\subsection*{Retrieve a phylogeny}
Ecologists are increasingly taking a phylogenetic approach to ecology, applying phylogenies to topics such as the study of community structure \cite{webb2002phylogenies}, ecological networks \cite{rafferty2013phylogenetic}, functional trait ecology \cite{poff2006functional}. Yet, Many biologists are not adequately trained in reconstructing phylogenies. Fortunately, there are some sources for getting a phylogeny without having to know how to build one; one of these is for angiosperms, called Phylomatic \cite{webb2005}. We have created a workflow in taxize that accepts a species list, and taxize works behind the scenes to get higher taxonomic names, which are required by Phylomatic to get a phylogeny. Here is a short example, producing the tree in figure \ref{fig:phylomatic}.

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{taxa} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"Poa annua"}\hlstd{,} \hlstr{"Abies procera"}\hlstd{,} \hlstr{"Helianthus annuus"}\hlstd{)}
\hlstd{tree} \hlkwb{<-} \hlkwd{phylomatic_tree}\hlstd{(}\hlkwc{taxa} \hlstd{= taxa)}
\hlstd{tree}\hlopt{$}\hlstd{tip.label} \hlkwb{<-} \hlkwd{capwords}\hlstd{(tree}\hlopt{$}\hlstd{tip.label)}
\hlkwd{plot}\hlstd{(tree,} \hlkwc{cex} \hlstd{=} \hlnum{1}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=3in]{phylogeny.pdf}
\end{center}
\caption{
{A phylogeny for three species. This phylogeny was produced using the \emph{phylomatic\_tree} function, which queries the Phylomatic database, and prunes a previously created phylogeny of plants.}
}
\label{fig:phylomatic}
\end{figure}

@@

\subsection*{What taxa are the children of my taxon of interest?}
If someone is not a taxonomic specialist on a particular taxon they probably do not know what children taxa are within a family, or within a genus. This task becomes especially unwieldy when there are a large number of taxa downstream. You can of course go to a website like Wikispecies \url{http://species.wikimedia.org/wiki/Main_Page} or Encyclopedia of Life \url{http://eol.org/} to get downstream names. However, taxize provides an easy way to programatically search for downstream taxa, both for the Catalogue of Life (CoL) \url{http://www.catalogueoflife.org/} and the Integrated Taxonomic Information System \url{http://www.itis.gov/}. Here is a short example using the CoL in which we want to find all the species within the genus \emph{Apis} (honey bees).

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{col_downstream}\hlstd{(}\hlkwc{name} \hlstd{=} \hlstr{"Apis"}\hlstd{,} \hlkwc{downto} \hlstd{=} \hlstr{"Species"}\hlstd{)[[}\hlnum{1}\hlstd{]]}
\end{alltt}
\end{kframe}
\end{knitrout}

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
  childtaxa_id     childtaxa_name childtaxa_rank
1      6971712 Apis andreniformis        Species
2      6971713        Apis cerana        Species
3      6971714       Apis dorsata        Species
4      6971715        Apis florea        Species
5      6971716 Apis koschevnikovi        Species
6      6845885     Apis mellifera        Species
7      6971717   Apis nigrocincta        Species
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\subsection*{Matching species tables with different taxonomic resolution}
Biologists often need to match different sets of data tied to species. For example, trait-based approaches are a promising tool in ecology \cite{statzner_can_2010}. One problem is that abundance data must be matched with trait databases such as the NCBI Taxonomy database \cite{usseglio-polatera_biological_2000}. These two data tables may contain species information on different taxonomic levels and data might have to be aggregated to a joint taxonomic level, so that the data can be merged. taxize can help in this data-cleaning step, providing a reproducible workflow.
